---
title: "Group Project: Data Exploration"
author: "Group X"
date: "10/28/2020"
output:
  word_document: default
  html_document: default
  pdf_document: default
---
  
```{r packages}
library(tidyverse)
```

# Download the csv file from the url:"https://www.kaggle.com/divyansh22/us-border-crossing-data/download" and unzip it before running.

``` {r}
Border_Crossing_Entry_Data <- read_csv("Border_Crossing_Entry_Data.csv")
View(Border_Crossing_Entry_Data)

write.csv(Border_Crossing_Entry_Data,"border.csv")
bdata <- Border_Crossing_Entry_Data
```
# Seeing a summary of original dataset
```{r}
head(bdata)
summary(bdata)
```
# Changing the variable name of the dataset so that it is easier to read. Also removing spaces from variable name if any.
```{r}
head(bdata)
cleaned_data <- bdata %>% 
  rename(Name = `Port Name`) %>% 
  rename(Code = `Port Code`) %>% 
  rename(Transportation = Measure) %>% 
  rename(Count = Value)

glimpse(cleaned_data)
```

# Rows with smaller count values (less than 50) were eliminated because of the negligible number of people entering as compared to other rows
```{r}
cleaned_data <- cleaned_data %>% 
  filter(Count > 50)
glimpse((cleaned_data))
```

# Verifying that there are two borders mentioned in the dataset
```{r}
head(cleaned_data %>%  group_by(Border) %>% summarise(Count = n()))
```

# Removing redundant information from _Border_ column
```{r}
cleaned_data <- cleaned_data %>% 
  
  mutate(Border = case_when(Border == "US-Canada Border" ~ "US-Canada", Border == "US-Mexico Border" ~ "US-Mexico"))
```
# Seeing a summary of cleaned dataset
```{r}
summary(cleaned_data)
```
# Any rows with missing values for variables were ignored. For example, if the value was empty
```{r}
cleaned_data <- na.omit(cleaned_data)
```


# Removing time value from _Date_ column because we don't need it as it gives no information besides making data look dirty
```{r}
cleaned_data <- cleaned_data %>% separate(Date, into = c("Date", "Time"), sep =" ")
cleaned_data <- cleaned_data %>% select(-Time)
head(cleaned_data)
```
# Keeping only Year information because we don't need Month and Day for the purpose of our study
```{r}
cleaned_data <- cleaned_data %>% separate(Date, into = c("Month", "Day", "Year"), sep ="/")
cleaned_data <- cleaned_data %>% select(-Month, -Day)
cleaned_data$Year <- as.numeric(cleaned_data$Year)
head(cleaned_data)
```



# How many borders are there
```{r}
unique(cleaned_data$State) %>% length()
```


# Which border has the most number of people coming in? 
```{r }
 cleaned_data %>% group_by(Border) %>% summarise(Total = sum(Count)) %>% ggplot() + geom_bar(mapping = aes(x=Border, y = Total, fill = Border), stat = "identity", width = 0.3) 
```
# Graph to see the pattern of number of people coming in from 1996-2020. We found less people are coming in recent years as compared to the past
```{r}
dummy <- cleaned_data %>% group_by(Border, Year) %>% summarise(Total = sum(Count), .groups = "keep")
ggplot(data = dummy, mapping=aes(x=Year, y=Total, color=Border)) + geom_point() + geom_line() + stat_smooth(method="lm")
```
# Graph to see the pattern of number of people coming in from 1996-2020 by state. We found less people are coming in recent years as compared to the past
```{r}
dummy <- cleaned_data %>% group_by(Border, Year, State) %>% summarise(Total = sum(Count), .groups = "keep")
ggplot(data = dummy, mapping=aes(x=Year, y=Total, color=State)) + geom_point() + geom_line() + stat_smooth(method="lm")
```

# Which State has the most people coming in?
```{r}
cleaned_data %>% group_by(Border, State) %>% summarise(Total = sum(Count), .groups = "keep") %>% arrange(desc(Total)) %>% ggplot() + geom_bar(mapping = aes(x=reorder(State, -Total), y = Total, color = State, fill = Border), stat = "identity") + labs(x = "State")
```

```{r}
head(cleaned_data)
unique(cleaned_data$Transportation)
```
# Looking at means of transportation by border
```{r}
cleaned_data %>% group_by(Border,Transportation) %>% summarise(Total = sum(Count), .groups = "keep") %>% arrange(desc(Total)) %>% ggplot() + geom_bar(mapping = aes(x=reorder(Transportation, -Total), y = Total, color =Transportation, fill = Border), position="fill", stat = "identity") + labs(x = "Means of Transportation") + coord_flip()
```

```{r}
cleaned_data %>% group_by(Border,Transportation) %>% summarise(Total = sum(Count), .groups = "keep") %>% arrange(desc(Total)) %>% ggplot() + geom_bar(mapping = aes(x=reorder(Transportation, -Total), y = Total, fill =Transportation),  stat = "identity") + labs(x = "Means of Transportation") + facet_wrap("Border") + coord_flip()
```

```{r}
cleaned_data %>% group_by(Border, State, Transportation) %>% summarise(Total=sum(Count), .groups = "keep") %>% ggplot() + geom_bar(mapping = aes(x=reorder(State, -Total), y=Total, fill =Transportation), position = "fill",  stat = "identity") + labs(x = "Means of Transportation") 
```
```{r}
cleaned_data %>% group_by(Border, State, Transportation) %>% summarise(Total=sum(Count), .groups = "keep") %>% ggplot() + geom_bar(mapping = aes(x=reorder(Transportation, -Total), y=Total, fill=Transportation),  stat = "identity") + labs(x = "Means of Transportation") + coord_flip() + facet_wrap("State")
```


# Multiple Regression Starts From Here
```{r}
df <- cleaned_data %>% group_by (Border,  Year, Transportation) %>% summarise(TotalEntry = sum(Count), .groups = "keep")
head(df)
```
```{r}
library(modelr)

library(cvTools)

fit.m<- lm(TotalEntry ~ Year + Transportation, data=df)
library(rpart)
library(rpart.plot)

fit.back <- step(fit.all, trace=0)
summary(fit.m)

df %>% add_predictions(fit.m, var="pred") %>% ggplot (mapping=aes(x=Year, y=TotalEntry, color=Transportation)) + geom_point() + geom_line(mapping=aes(x=Year, y=pred, color=Transportation))

df %>% add_predictions(fit.m, var="pred") %>% add_residuals(fit.m, var="resd") %>% ggplot (mapping=aes(x=Year, y=resd, color=Transportation)) + geom_point() 

unique(df$Transportation)
```
# CART
```{r}
df <- df %>% mutate(Transportation = case_when(Transportation == "Bus Passengers" ~ "Bus", 
                                         Transportation == "Buses" ~ "Bus", 
                                         Transportation == "Pedestrians" ~ "Pedestrians", Transportation == "Personal Vehicle Passengers" ~ "Personal Vehicle",
Transportation == "Personal Vehicles " ~ "Personal Vehicle", 
Transportation == "Rail Containers Empty" ~ "Rail", 
Transportation == "Rail Containers Full" ~ "Rail", 
Transportation == "Train Passengers" ~ "Train", 
Transportation == "Trains" ~ "Train", 
Transportation == "Truck Containers Full" ~ "Truck", 
Transportation == "Truck Container Empty" ~ "Truck", 
Transportation == "Trucks" ~ "Truck"))

unique(df$Transportation)


```
```{r}
fit.rpart <- rpart(TotalEntry ~ Year + Transportation, data=df)
fit.rpart
rpart.plot(fit.rpart)
prp(fit.rpart)
```


# Saving the cleaned data file
```{r}
write.csv(cleaned_data, "us_border_cleaned.csv")
```

